Museum specimens as baseline for assessing a potential global pollinator crisis

### 2018-03-23 ms.Rmd
updtaed in text citations. Also added reference file and citation style file (Phil B).

---
title: "Museum specimens as baseline for assessing a potential global pollinator crisis"
author: "I. Bartomeus, J. Stavert, D. Ward, O. Aguado"
output: html_document
---

@@

Animal pollinators are a critical component of both natural and production ecosystems worldwide, given their role in plant reproduction (@ollerton2012) and food security (@kleinetal2007). As with many other taxa, pollinators are vulnerable to a range of anthropogenic disturbances, which can cause local and regional population declines or even extinctions. This issue was identified several decades ago, and was popularized in 1996 by the influential book "the forgotten pollinators". However, these early accounts were somewhat anecdotal, given the lack of data on pollinator populations at that time. These initial claims triggered the first efforts to asses the problem, including the formation of a US National Academy of Science (NAS) panel in 2006, which was commissioned to assess the extent of pollinator declines. The NAS report concluded that "For most pollinator species [...] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult". Since then, studies on pollinator responses to various global change drivers have rapidly multiplied and researchers have developed a strong consensus that disturbances such as habitat destruction, land-use change, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically (@goulson2016, @potts2010). Yet the current status and population trends of most pollinator species worldwide remain unknown. For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, 60% of bee species [check number]() fell into the "Data Deficient" category (@Nieto2014). For countries outside of Europe and the US, data on pollinator populations is almost non-existent.

One of the main barriers to identifying trends in pollinator populations is that pollinators are incredibly diverse and including bees, flies, butterflies, beetles, birds, bats and even lizards (@Winfreeetal2011). Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution (@ollerton2018). However, bees are also extremely diverse, with more than 20,000 species currently described worldwide (@PickeringAscher), and require often require expert taxonomists for identification. The uneven distribution of bee researchers has resulted in geographical biases in bee decline research (@Martinetal2012) and there are also taxonomic biases toward species that are easier to identify, such bumblebees (@ArcherOikos). Another barrier to identifying pollinator declines is the difficulty in assessing population trends for these often highly mobile, short-lived and small organisms.
One solution to overcoming these barriers has been to use space-for-time substitutions, where researchers compare pollinator populations across environmental gradients. Despite critiques on the robustness of this approach (@ref), this is the most extensive source of data on pollinator populations currently available. In Europe, researchers have recently estimated bee richness declines for every country using predictions generated from models [using what parameters?] (@DePalma). A second promising method is assessing data collected in pollinator monitoring programs, which are often driven by citizen scientists. This approach was inspired by successful butterfly monitoring programs in countries such as the UK and is currently being extended to other pollinator taxa (@ref). However, these programs require significant time to generate long-term datasets and cannot reveal anything about historic pollinator populations. Finally, the most robust approach for assessing long-term pollinator population trends is to use museum specimen collection data. The core aim of museums is to conserve and curate collections. Thus, they serve as a precious repository for specimens and also ensure high quality taxonomic identification (@primak). Yet currently, the major bottleneck for researchers wanting to use contained within museum collections is the digitization of collection data. Where digitization has been completed (e.g., US and Europe), the data provide a rich source of information and have allowed the assessment of the current statuses and long-term trends in pollinator populations (e.g. Bartomeus et al 2013, Scheper et al 2014). This is despite the fact that museum collections often have a number of biases [including...] (@Keil). But in many cases, museum collections are the only data we have to reconstruct the past.

Regardless of the method, it is important for researchers to consider aspects of biodiversity beyond species richness. In addition to local (alpha) diversity and regional (gamma) diversity, we need to measure changes in turnover between sites (beta diversity), abundance of key species, and evenness of communities (@winfree2016ELE). Environmental changes often results in few “winners” and many “losers” (@Bartomeusetal2013). Identifying winners and losers is critical as the few winner species are often exotic and represent a subset of traits that facilitate survival in highly modified environments (@Stavertetal2018). These changes can have important flow on effects for pollination of native plant species and crops.  [Add something about traits of winners and losers?]()
https://www.sciencedirect.com/science/article/pii/S0065250417300296
[move this and expand to the discussion on the map?]()

@@

Digitizing old specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification (@ref cory paper on taxonomy needs), (ii) geo-locate the coordinates of collection events (@ref and software) and (iii) store the data in a properly curated database (@ref on database management). Undertaking this process for the thousands of specimens in museum collections can be a daunting task and requires specialized personnel. While some tasks can only be undertaken by people with specialist skills (e.g., taxonomists), new technologies and citizen science can speed up the collection digitization process. High resolution specimen photos and associated labels can be uploaded to the internet, where the job of image transcription can be distributed across hundreds or thousands of volunteers. In addition to this, new algorithms have been created that allow location geo-referencing based on vernacular names (@refs geonames, geoparser and opencage). [Explain zooniverse, NZ initiative and how data can be validated... Is there anything on AI to automate the process? e.g. there is something on wing venation from photos at Netherlands. Can be mentioned?]

@@

It is unsurprising that most evidence of pollinator declines is biased towards developed western countries, which have also been subject to extensive anthropogenic disturbance. For example, in the UK and the Netherlands, which have experienced a multitude of widespread anthropogenic changes, a citizen science based study using and museum collections detected declines for bees, flies and flowering plants (@Beisjmieretal2006). However, it is important to note that even for these two countries, pollinator richness local estimates are biased toward large cities and regions dominated by agriculture, and data is lacking for natural areas. Further exploration of this dataset revealed that the declining trend for pollinator taxa has attenuated in recent decades (@carvalheiroetal). More detailed trends for common bee species in the Netherlands have also been explored using museum specimens and confirm the link between simultaneous plant and pollinator declines (@Schepper2014). This study found that bee species with the strongest host plant preferences (i.e., specialists) displayed the strongest declines and thus were most threatened with extinction. [check this] [- EU: UK? Ollerton Science, Peter Keil flies? Moths? Buterflies! Add Ollerton paper!]()

In comparison to local decline estimates, regional estimates for changes in species richness in the eastern US show a moderate decline in bee diversity (@Bartomeusetal2013) and very few regional extinctions (@Colla2012?). These findings are in stark contrast with the widespread local extinctions reported by Burkle (@Burkle), who compared historical observations of bee occurrence in a large forested ecosystem with the present-day remaining forest remnants. However, it is important to note that this comparison does not take into account opportunistic bee species that use novel land-use types. Thus, studies may detect local richness declines, while regional richness remains relatively stable. In any case, there is strong concordance between local extinctions and regional declines (@Bartomeus_winfree2013), suggesting that local extinctions are indicators of regional population declines.

@@

Museum can provide valuable information besides species occurrence records, and this is important to take into account when digitizing the collections. This is specially relevant to identify mechanisms of decline and adaptation. One particularly important source of information is the date of capture, which has allowed to track phenological advances congruent with contemporary climate change (@Bartomeusetal2011). In addition, as mentioned above, insect pollinator labels often include as information the host plant where collected. This information is precious to understand past and present species interactions (@tylianakis2010ELE). Even when the host plant is not noted, bee specimens often contain pollen loads trapped in the hairs, making possible to reconstruct past visitation events (@kleinj_raemakers). Finally, museum specimens can be measured for specific attributes and this has been done to show tongue length (@refAlaskaScience) and body size (@Oliveira) shrinkage with climate and land use change.   
[move this section below]()

@@

Gbif (@ref) is a central repository centralizing occurrence data. Much of this data comes from museum, but it integrates several other sources. Alongside the popular statistical language R (@ref), gbif can be directly queried into your computer (@rgbif) and check data availability for your region of interest. In Figure 2 we show the number of modern and historical bee records already available per country. While for some countries like UK and US the numbers are high in both axes, and hence, data can be potentially analyzed without further data collection efforts, most countries fall short in one or both of the axes. Hence, for countries with large number of recent records, researchers should prioritize the digitalization of old material to conduct their analysis. On the contrary, for countries with large number of historical records, organized re-surveys should be organized (e.g. @cameron2013). Interestingly, we show that >100 countries have less than 1000 records, making them unlikely candidates to study, but countries like X or X... Also note that historical records are not always in local museums, but many UK and US museums contain large collection of exotic pollinators.

@@

##Case study two: NZ    
In contrast to Spain, Aotearoa-New Zealand is an isolated oceanic archipelago, with a distinctive pollinator biota and a unique history of human occupation. Much of Aotearoa-New Zealand’s pollinator fauna is also relatively depauperate. For example, Aotearoa-New Zealand has only 27 species of native bees (@Donovan, 2007), which is a fraction of nearby Australia’s c. 1600 species (REF). However, Aotearoa-New Zealand has a surprisingly high diversity of fly (Diptera) pollinators, which are important pollinators in many ecosystems (@Newstrom & Robertson, 2005). Thus, Aotearoa-New Zealand provides a unique system to study long-term changes in pollinator communities, and is unlike continental Europe and the US, which have been the focus of an overwhelming majority of pollinator decline studies.

In global terms, human colonisation of Aotearoa-New Zealand was relatively recent (c. 740 y) (@Wilmshurst et al. 2008). Before human arrival, Aotearoa-New Zealand was predominately forested, but has since been dramatically altered by people. Early Māori settlers cleared forests by burning and more recently, European colonists cleared large tracts of remaining forests and drained low-lying wetlands for agriculture, mostly before 1900 (@Perry et al, 2014). Therefore, human activity likely affected pollinator communities in Aotearoa-New Zealand long before surveys and collections began. Nevertheless, we can use museum records to identify trends in pollinator communities during Aotearoa-New Zealand’s more recent history.

@@

Data analyses
First, we filtered our original datasets so that data used for analyses only included unique collection events. To do this, we removed specimens collected at the same location, on the same date, and by the same collector. We found our data had reasonable coverage across time periods, although there was a peak in collection occurrences from 1960-1980. We accounted for differences in collection effort through binning collection records by time so that each bin had a similar number of records but a different number of years (see @Bartomeusetal2013 for further details). We then estimated richness for each time period bin by rarefying all bins to an equal number of specimens and calculated the mean species richness ±SE for each bin. Finally, we estimated the significance of change in richness using a permutation test that randomly reordered time periods and calculated the correlation between time period and species richness. Thus reported P-values were the proportion of permutations that had higher or lower correlations compared to the correlation between richness and the actual chronological time period sequence.

@@

As illustrated in our examples, unleashing the power of museum collection data to answer pressing questions is at our hands, but requires the coordinated effort of many actors. Using case studies, we show that collaboration with both museum curators and data collectors is key to adequately understanding the data and treating it appropriately. We envision that instead of a central museum collection record repository, researchers and curators should simply aim to make digitalized data readily available and easy to share. Ultimately, centralisation of these data in existing platforms such as Gbif would facilitate free and widespread access. However, even when data are stored in different webpages or databases, it can be retried and combined with other datasets using open science tools (@ropensci), providing these tools are accessible (API e.g. @heltmint).

@@

Ultimately, to identify global trends in pollinator declines we require robust data, collected from diverse geographic regions. Further, it is crucial that data are analysed in the appropriate way. This requires researches identifying biases and filling in gaps where possible. We need to put increased emphasis on quantifying pollinator declines in regions outside of the US and Europe and for pollinator groups other than bees. For the US and Europe there have been few regional bee extinctions (@Ollerton, @Bartomeus2013), but declines are widespread in disturbed ecosystems (@Bartomeus, @dePalma, @Carvalheiro). For most other pollinator taxa and regions throughout the world we know almost nothing. Moving forward, the first step for many taxa will be to identify and describe species. Only then can we begin to document declines.

@@

>>>>

---
title: "Museum specimens as baseline for assessing a potential global pollinator crisis"
author: "I. Bartomeus, J. Stavert, D. Ward, O. Aguado"
output: html_document
bibliography: data/references.xml
csl: data/philb_style.xml
---

@@

Animal pollinators are a critical component of both natural and production ecosystems worldwide, given their role in plant reproduction [#@Ollerton2012] and food security [@Klein2007]. As with many other taxa, pollinators are vulnerable to a range of anthropogenic disturbances, which can cause local and regional population declines or even extinctions. This issue was identified several decades ago, and was popularized in 1996 by the influential book "the forgotten pollinators". However, these early accounts were somewhat anecdotal, given the lack of data on pollinator populations at that time. These initial claims triggered the first efforts to asses the problem, including the formation of a US National Academy of Science (NAS) panel in 2006, which was commissioned to assess the extent of pollinator declines. The NAS report concluded that "For most pollinator species [...] the paucity of long-term population data and the incomplete knowledge of even basic taxonomy and ecology make definitive assessment of status exceedingly difficult". Since then, studies on pollinator responses to various global change drivers have rapidly multiplied and researchers have developed a strong consensus that disturbances such as habitat destruction, land-use change, chemical exposure, exotic species and climate change are causing pollinator declines, and often act synergistically [@Goulson2015; @Potts2010]. Yet the current status and population trends of most pollinator species worldwide remain unknown. For example, a recent IUCN report concluded that even for Europe’s comparatively well-studied bee fauna, 60% of bee species [check number]() fell into the "Data Deficient" category [@Nieto2014]. For countries outside of Europe and the US, data on pollinator populations is almost non-existent.

One of the main barriers to identifying trends in pollinator populations is that pollinators are incredibly diverse and including bees, flies, butterflies, beetles, birds, bats and even lizards [#@Winfree2011]. Bees are generally regarded as the most important pollinator group due to their abundance, pollination efficiency and widespread distribution [@Ollerton2017]. However, bees are also extremely diverse, with more than 20,000 species currently described worldwide [#@PickeringAscher], and require often require expert taxonomists for identification. The uneven distribution of bee researchers has resulted in geographical biases in bee decline research [#@Martinetal2012] and there are also taxonomic biases toward species that are easier to identify, such bumblebees [#@ArcherOikos]. Another barrier to identifying pollinator declines is the difficulty in assessing population trends for these often highly mobile, short-lived and small organisms.
One solution to overcoming these barriers has been to use space-for-time substitutions, where researchers compare pollinator populations across environmental gradients. Despite critiques on the robustness of this approach [@ref], this is the most extensive source of data on pollinator populations currently available. In Europe, researchers have recently estimated bee richness declines for every country using predictions generated from models (using what parameters?) [@DePalma2017]. A second promising method is assessing data collected in pollinator monitoring programs, which are often driven by citizen scientists. This approach was inspired by successful butterfly monitoring programs in countries such as the UK and is currently being extended to other pollinator taxa [@ref]. However, these programs require significant time to generate long-term datasets and cannot reveal anything about historic pollinator populations. Finally, the most robust approach for assessing long-term pollinator population trends is to use museum specimen collection data. The core aim of museums is to conserve and curate collections. Thus, they serve as a precious repository for specimens and also ensure high quality taxonomic identification [@primak]. Yet currently, the major bottleneck for researchers wanting to use contained within museum collections is the digitization of collection data. Where digitization has been completed (e.g., US and Europe), the data provide a rich source of information and have allowed the assessment of the current statuses and long-term trends in pollinator populations (e.g., @Bartomeus2013a; #@Scheper2014). This is despite the fact that museum collections often have a number of biases [including...] [@Keil]. But in many cases, museum collections are the only data we have to reconstruct the past.

Regardless of the method, it is important for researchers to consider aspects of biodiversity beyond species richness. In addition to local (alpha) diversity and regional (gamma) diversity, we need to measure changes in turnover between sites (beta diversity), abundance of key species, and evenness of communities [@Winfree2015]. Environmental changes often results in few “winners” and many “losers” [@Bartomeus2013b]. Identifying winners and losers is critical as the few winner species are often exotic and represent a subset of traits that facilitate survival in highly modified environments [@Stavert2018]. These changes can have important flow on effects for pollination of native plant species and crops.  [Add something about traits of winners and losers?]()
https://www.sciencedirect.com/science/article/pii/S0065250417300296
[move this and expand to the discussion on the map?]()

@@

Digitizing old specimens is not a trivial task and requires expertise to (i) ensure proper taxonomic identification [@ref cory paper on taxonomy needs], (ii) geo-locate the coordinates of collection events [@ref and software] and (iii) store the data in a properly curated database [@ref on database management]. Undertaking this process for the thousands of specimens in museum collections can be a daunting task and requires specialized personnel. While some tasks can only be undertaken by people with specialist skills (e.g., taxonomists), new technologies and citizen science can speed up the collection digitization process. High resolution specimen photos and associated labels can be uploaded to the internet, where the job of image transcription can be distributed across hundreds or thousands of volunteers. In addition to this, new algorithms have been created that allow location geo-referencing based on vernacular names [@refs geonames, geoparser and opencage]. [Explain zooniverse, NZ initiative and how data can be validated... Is there anything on AI to automate the process? e.g. there is something on wing venation from photos at Netherlands. Can be mentioned?]

@@

It is unsurprising that most evidence of pollinator declines is biased towards developed western countries, which have also been subject to extensive anthropogenic disturbance. For example, in the UK and the Netherlands, which have experienced a multitude of widespread anthropogenic changes, a citizen science based study using and museum collections detected declines for bees, flies and flowering plants [@Biesmeijer2006]. However, it is important to note that even for these two countries, pollinator richness local estimates are biased toward large cities and regions dominated by agriculture, and data is lacking for natural areas. Further exploration of this dataset revealed that the declining trend for pollinator taxa has attenuated in recent decades [@Carvalheiro2013]. More detailed trends for common bee species in the Netherlands have also been explored using museum specimens and confirm the link between simultaneous plant and pollinator declines [#@Scheper2014]. This study found that bee species with the strongest host plant preferences (i.e., specialists) displayed the strongest declines and thus were most threatened with extinction. [check this] [- EU: UK? Ollerton Science, Peter Keil flies? Moths? Buterflies! Add Ollerton paper!]()

In comparison to local decline estimates, regional estimates for changes in species richness in the eastern US show a moderate decline in bee diversity [@Bartomeus2013a] and very few regional extinctions [@Colla2012]. These findings are in stark contrast with the widespread local extinctions reported by Burkle [@Burkle2013], who compared historical observations of bee occurrence in a large forested ecosystem with the present-day remaining forest remnants. However, it is important to note that this comparison does not take into account opportunistic bee species that use novel land-use types. Thus, studies may detect local richness declines, while regional richness remains relatively stable. In any case, there is strong concordance between local extinctions and regional declines [#@Bartomeus_Winfree2013], suggesting that local extinctions are indicators of regional population declines.

@@

Museum can provide valuable information besides species occurrence records, and this is important to take into account when digitizing the collections. This is specially relevant to identify mechanisms of decline and adaptation. One particularly important source of information is the date of capture, which has allowed to track phenological advances congruent with contemporary climate change [#@Bartomeus2011]. In addition, as mentioned above, insect pollinator labels often include as information the host plant where collected. This information is precious to understand past and present species interactions [#@Tylianakis2010ELE]. Even when the host plant is not noted, bee specimens often contain pollen loads trapped in the hairs, making possible to reconstruct past visitation events [#@kleinj_raemakers]. Finally, museum specimens can be measured for specific attributes and this has been done to show tongue length [#@refAlaskaScience] and body size [#@Oliveira] shrinkage with climate and land use change.   
[move this section below]()

@@

Gbif [@ref] is a central repository centralizing occurrence data. Much of this data comes from museum, but it integrates several other sources. Alongside the popular statistical language R [@ref], gbif can be directly queried into your computer [#@rgbif] and check data availability for your region of interest. In Figure 2 we show the number of modern and historical bee records already available per country. While for some countries like UK and US the numbers are high in both axes, and hence, data can be potentially analyzed without further data collection efforts, most countries fall short in one or both of the axes. Hence, for countries with large number of recent records, researchers should prioritize the digitalization of old material to conduct their analysis. On the contrary, for countries with large number of historical records, organized re-surveys should be organized [e.g. #@cameron2013]. Interestingly, we show that >100 countries have less than 1000 records, making them unlikely candidates to study, but countries like X or X... Also note that historical records are not always in local museums, but many UK and US museums contain large collection of exotic pollinators.

@@

##Case study two: NZ    
In contrast to Spain, Aotearoa-New Zealand is an isolated oceanic archipelago, with a distinctive pollinator biota and a unique history of human occupation. Much of Aotearoa-New Zealand’s pollinator fauna is also relatively depauperate. For example, Aotearoa-New Zealand has only 27 species of native bees [@Donovan2007], which is a fraction of nearby Australia’s c. 1600 species [@REF]. However, Aotearoa-New Zealand has a surprisingly high diversity of fly (Diptera) pollinators, which are important pollinators in many ecosystems [@Newstrom2005]. Thus, Aotearoa-New Zealand provides a unique system to study long-term changes in pollinator communities, and is unlike continental Europe and the US, which have been the focus of an overwhelming majority of pollinator decline studies.

In global terms, human colonisation of Aotearoa-New Zealand was relatively recent (c. 740 y) [@Wilmshurst2008]. Before human arrival, Aotearoa-New Zealand was predominately forested, but has since been dramatically altered by people. Early Māori settlers cleared forests by burning and more recently, European colonists cleared large tracts of remaining forests and drained low-lying wetlands for agriculture, mostly before 1900 [@Perry2014]. Therefore, human activity likely affected pollinator communities in Aotearoa-New Zealand long before surveys and collections began. Nevertheless, we can use museum records to identify trends in pollinator communities during Aotearoa-New Zealand’s more recent history.

@@

Data analyses
First, we filtered our original datasets so that data used for analyses only included unique collection events. To do this, we removed specimens collected at the same location, on the same date, and by the same collector. We found our data had reasonable coverage across time periods, although there was a peak in collection occurrences from 1960-1980. We accounted for differences in collection effort through binning collection records by time so that each bin had a similar number of records but a different number of years [see @Bartomeus2013a for further details]. We then estimated richness for each time period bin by rarefying all bins to an equal number of specimens and calculated the mean species richness ±SE for each bin. Finally, we estimated the significance of change in richness using a permutation test that randomly reordered time periods and calculated the correlation between time period and species richness. Thus reported P-values were the proportion of permutations that had higher or lower correlations compared to the correlation between richness and the actual chronological time period sequence.

@@

As illustrated in our examples, unleashing the power of museum collection data to answer pressing questions is at our hands, but requires the coordinated effort of many actors. Using case studies, we show that collaboration with both museum curators and data collectors is key to adequately understanding the data and treating it appropriately. We envision that instead of a central museum collection record repository, researchers and curators should simply aim to make digitalized data readily available and easy to share. Ultimately, centralisation of these data in existing platforms such as Gbif would facilitate free and widespread access. However, even when data are stored in different webpages or databases, it can be retried and combined with other datasets using open science tools (#@ropensci), providing these tools are accessible [API e.g., #@heltmint].

@@

Ultimately, to identify global trends in pollinator declines we require robust data, collected from diverse geographic regions. Further, it is crucial that data are analysed in the appropriate way. This requires researches identifying biases and filling in gaps where possible. We need to put increased emphasis on quantifying pollinator declines in regions outside of the US and Europe and for pollinator groups other than bees. For the US and Europe there have been few regional bee extinctions [@Ollerton; @Bartomeus2013c] (#check this Bartomeus reference), but declines are widespread in disturbed ecosystems [@Bartomeus2013a; @DePalma2017; @Carvalheiro2013]. For most other pollinator taxa and regions throughout the world we know almost nothing. Moving forward, the first step for many taxa will be to identify and describe species. Only then can we begin to document declines.

@@

#References
